Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-30 index.qmd
section:background,dag; process:write,images

  Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs of stimuli to determine which one exhibits a higher degree of a specific trait. DCJ has proven effective and reliable across various fields [@Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020]. However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data could be generated. This study addresses these issues by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach, the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system. It then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal effects within the system. This research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927]. Its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.

@@

@@

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify the causal assumptions and mechanisms inherent in the DCJ system. Next, using a minimal set of assumptions, the study translates the scientific model into a probabilistic statistical model. This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.

@@

# Theoretical background {#sec-background}

## The structural approach to causal inference {#sec-background-structural}

@@

<!-- 2. what are counterfactuals? -->
Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. According to @Pearl_et_al_2018, counterfactuals form the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association. Despite this abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018]. They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances."

@@

## DAGs, SCMs, and the flow of association and causation in graphs {#sec-background-dag}

@@

<!-- 2. How graphs are use in the structural approach? and what are DAGs? -->
In statistics, one application that incorporates concepts from graph theory is causal inference. Specifically, the structural approach to causal inference uses DAGs to provide a formal and graphical representation of a system's causal structure [@Neal_2020]. In this context, a graph is a collection of nodes connected by edges, where nodes represent random variables. In a directed graph, edges extend from one node to another, with arrows indicating the direction of causal influence. In a directed acyclic graph, the direction of causal influences does not form a loop, ensuring that causal influences do not cycle back on themselves [@McElreath_2020].

<!-- 8. What are the DAG's building blocks?
<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->

<!-- 4. A motivating example -->
Given the heuristic nature of DAGs, these are easier to understand using a motivating example. Consider a system where variables $X$ and $Z$ influence a third variable $Y$. In this system, it is assumed that $X$ and $Z$ are dependent on their own processes and are, therefore, independent from each other. @fig-dag1a presents the plausible causal structure of this system. The DAG shows the endogenous variables $V=\{X,Z,Y\}$ as circled black nodes, indicating these variables are observed. The arrows connecting the variables indicate the direction of causal influence, while a lack of influence is indicated by the absence of arrows. Moreover, the exogenous variables $E=\{e_{X},e_{Z},e_{Y}\}$ represent everything else that is chosen not to be modeled explicitly. These exogenous variables, or *disturbances*, are usually represented by open circles to indicate their unobserved nature. Although this DAG explicitly shows the exogenous variables, conventionally these are omitted for brevity, resulting in an equivalent graph as shown in @fig-dag1b [@Cinelli_et_al_2020].

@@

::: {#fig-dags layout-ncol=2}

![Full DAG](images/figures/dag1a.png){#fig-dag1a fig-align="center" width=90%}

![Simplified DAG](images/figures/dag1b.png){#fig-dag1b fig-align="center" width=35%}

@@

@@

@@

# Theoretical framework {#sec-theory}

>>>>

  Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs of stimuli to determine which one exhibits a higher degree of a specific trait. DCJ has proven effective and reliable across various fields [@Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020]. However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data could be generated. This study addresses these issues by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system. The study then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal effects within the system. This research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927]. Its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.

@@

lightbox: true

@@

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability [@Bramley_2015; @Pollitt_2012b; @Bramley_et_al_2019; @Verhavert_et_al_2019; @Crompvoets_et_al_2022; @vanDaal_et_al_2017; @Gijsen_et_al_2021], none, to the best of the authors' knowledge, have provided a transparent depiction of the DCJ system and the mechanisms generating the DCJ outcome. This study aims to fill this gap by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach [@Wright_1921; @Pearl_2009; @Pearl_et_al_2016], the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system. The study then translates the scientific model into a probabilistic statistical model. This model aims to produce statistical estimates to draw inferences about plausible causal relationships within the DCJ system.

@@

# Theoretical framework {#sec-framework}

## The structural approach to causal inference {#sec-framework-structural}

@@

<!-- 2. what are counterfactuals? -->
Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. According to @Pearl_et_al_2018, counterfactuals form the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association. Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018]. They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances."

@@

## Directed acyclic graphs (DAG) {#sec-framework-dag}

@@

<!-- 2. How graphs are use in the structural approach? and what are DAGs? -->
In statistics, one application that incorporates concepts from graph theory is causal inference. Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a formal and graphical representation of a system's causal structure [@Neal_2020]. In this context, a *graph* is a collection of nodes connected by edges, where nodes represent random variables. The term *directed* indicates that the edges of the graph extend from a *parent* node to a *child* node, with arrows showing the direction of causal influence. The term *acyclic* indicates that the direction of causal influences does not form a loop, ensuring that the influences do not cycle back on themselves [@McElreath_2020].

<!-- 3. How do we draw DAGs? -->
A key advantage of DAGs is that, regardless of the complexity of the causal structure, these structures can be composed using a minimal set of building blocks. Moreover, representing any complex structure requires only five fundamental building blocks [@Neal_2020; @McElreath_2020]. These advantages imply that a researcher seeking to depict and understand the causal assumptions of a complex system only needs to decompose the structure into these basic building blocks and then analyze the causal assumptions associated with each block [@McElreath_2020].

<!-- 4. What are DAG's building blocks?  -->
@fig-dags presents the five fundamental building blocks of any DAG. @fig-dag_bb1 illustrates two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related. @fig-dag_bb2 shows two connected nodes, depicting an scenario where a *parent* node $X_{1}$ has a causal influence on a *child* node $X_{2}$. This implies that variable $X_{2}$ is a *descendant* of $X_{1}$. @fig-dag_bb3 illustrates a *chain* (or *pipe*), where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$. This DAG indicates that $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ is a parent node $X_{3}$. Furthermore, it shows that $X_{1}$ is an *ancestor* of $X_{3}$, with the relationship between the two variables being mediated entirely by $X_{2}$. @fig-dag_bb4 depicts a *fork*, illustrating an scenario where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$. In this scenario $X_{2}$ is a parent node for both $X_{1}$ and $X_{3}$. Finally, @fig-dag_bb5 illustrates a *collider* (also known as *inmorality*), where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$. In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$. Given the heuristic nature of DAGs, the use of fundamental building block to construct a causal structure of a system is easier to understand using a motivating example. The motivating example can also serve to point out about other conventions when using DAGs.

::: {#fig-dags layout="[[1,1],[1,1,1]]"}

![Two unconnected nodes](images/figures/dag_bb1.png){#fig-dag_bb1 fig-align="center" width=40%}

![Two connected nodes or descendant](images/figures/dag_bb2.png){#fig-dag_bb2 fig-align="center" width=40%}

![Chain or pipe](images/figures/dag_bb3.png){#fig-dag_bb3 fig-align="center" width=100%}

![Fork](images/figures/dag_bb4.png){#fig-dag_bb4 fig-align="center" width=100%}

![Collider or inmorality](images/figures/dag_bb5.png){#fig-dag_bb5 fig-align="center" width=100%}

DAG's fundamental building blocks.
:::

<!-- 4. A motivating example -->
Consider a system where variables $X$ and $Z$ influence a third variable $Y$. In this system, it is assumed that $X$ and $Z$ are dependent on their own processes and are, therefore, independent from each other. @fig-dag_example1 presents the plausible causal structure of this system. The DAG shows the endogenous variables $V=\{X,Z,Y\}$ as circled black nodes, indicating these variables are observed. The arrows connecting the variables indicate the direction of causal influence, while a lack of influence is indicated by the absence of arrows. Moreover, the exogenous variables $E=\{e_{X},e_{Z},e_{Y}\}$ represent everything else that is chosen not to be modeled explicitly. These exogenous variables, or *disturbances*, are usually represented by open circles to indicate their unobserved nature. Although this DAG explicitly shows the exogenous variables, conventionally these are omitted for brevity, resulting in an equivalent graph as shown in @fig-dag_example2 [@Cinelli_et_al_2020].

@@

::: {#fig-example layout-ncol=2}

![Full DAG](images/figures/dag_example1.png){#fig-dag_example1 fig-align="center" width=100%}

![Simplified DAG](images/figures/dag_example2.png){#fig-dag_example2 fig-align="center" width=40%}

@@

## The flow of association and causation in graphs {#sec-framework-flow}

@@

<!-- Ultimately, DAGs play the important role of identifying which variables should be controlled for to be able to estimate causal effects. -->

@@

# Theory {#sec-theory}
